High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-04-03 PNAS_manuscript/PNAS_manuscript.tex

\affil[a]{Department of Earth and Planetary Science, University of California, Berkeley, CA, 94720}
\affil[b]{U.S. Geological Survey, Moffett Field, CA, 94025}
\affil[c]{Department of Earth and Planetary Sciences, Harvard University, Cambridge, MA, 02138}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy \cite{Buffett2000a}. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates have implied higher thermal conductivity values and an age for the inner core that is less than 1.5 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b, Frost2022a}, with some suggesting even younger ages ($<$700 Ma;  \citealp{Labrosse2015a, Ohta2016a}). Given that estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of Earth's core and mantle.

@@

Determinations of the absolute value of ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain a record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluids. However, intraplate magmatism associated with large igneous provinces is typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift (Fig. \ref{fig:Geologic_map}) is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma during which there were pulsed intervals of more rapid activity \cite{Swanson-Hysell2021a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and paleomagnetic data from rift rocks forms a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it enables an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

@@

In this study, we target a unique rock type---high-purity anorthosite xenoliths. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud magnetic inclusions from alteration with alteration of the plagioclase crystals themselves not resulting in the formation of secondary iron oxides in contrast with mafic minerals such as olivine. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They cooled and acquired their magnetizations in conjunction with the host diabase at a paleolatitude of 22\textdegree $\pm$2\textdegree (calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a, Zhang2021b}). Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows that the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals without strong preferred orientations. These single-slope, high-quality paleointensity data, together with other high-quality paleointensity records during the protracted Midcontinent Rift magmatic activity, require that there was a strong late Mesoproterozoic geodynamo.

@@

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths to have dominantly single-component magnetization that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with magnetization held by low-titanium titanomagnetite \cite{Zhang2021b}.

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). Plagioclase crystals within coarse-grained intrusions often contain abundant elongate Fe-Ti oxide inclusions that are visible through optical microscopy \cite{Feinberg2006a}. The Beaver River anorthosite xenoliths lack such large (10s of $\mu$m) oxide inclusions such that oxides are not visible within the plagioclase crystals through optical microscopy (Fig. \ref{fig:Petro_QDM}C). However, magnetic imaging using a quantum diamond microscope (QDM) shows that there are magnetic remanence carriers within the plagioclase crystals (Fig. \ref{fig:Petro_QDM}E, G). To highlight these distinct aspects of the Beaver River anorthosite xenoliths, we present petrographic and magnetic imaging data from the Beaver River anorthosite as well as a sample of an anorthosite from the Duluth Complex Anorthositic Series---an older intrusive complex within the Midcontinent Rift that was not targeted for paleointensity experiments in this study (Fig. \ref{fig:Petro_QDM}). In contrast to the Beaver River anorthosite, the plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display pronounced igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles within the Duluth Complex plagioclase grains that are typically aligned with the [001] axes of the crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these magnetite needles have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}F, H). As a result of this shape anisotropy, magnetization is acquired at angles highly oblique to applied fields (Figure \ref{fig:Petro_QDM}F). In an experiment where a second orthogonal field was applied, the first isothermal remanent magnetizations (IRM) of a set of magnetite needles within a plagioclase grain either fail to rotate or flip by 180\textdegree. When they flip, they remain in a direction that is oblique to the applied field direction dictated by shape anisotropy (Fig. \ref{fig:Petro_QDM}F, H). These experiments enable novel visualization of the grain-scale magnetic anisotropy of elongated exsolved (titano)magnetite inclusions within plagioclase that leads to magnetic anisotropy observed at the bulk sample scale \cite{Selkin2000a, Feinberg2006a}. In contrast, the remanent magnetizations of the ferromagnetic grains imaged in the Beaver River anorthosite xenoliths, targeted for paleointensity in this study, align with the applied field directions indicating minimal remanence anisotropy  (Fig. \ref{fig:Petro_QDM}E, G). The relative lack of petrofabrics and minimal grain-scale magnetic anisotropy make the Beaver River anorthosite xenoliths a particularly compelling target for paleointensity experiments.

@@

Following IZZI-style paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase sites have specimen results that pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1) where each site is an individual anorthosite xenolith. The paleointensity quality index (Q$_{PI}$; \citealp{Biggin2014a}) for the anorthosite xenolith sites are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from the anorthosite xenoliths have a mean of 38.28$\pm$11.92 $\mu$T. The site-mean virtual dipole moment is $\sim$82 ZAm$^2$ ($10^{21} Am^2$) ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthorg/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments are displayed on Arai plots and zero-field heating results are shown on inset orthogonal plots (Zijderveld plots) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent magnetization (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent fits for the dominant single-slope component that passes the acceptance criteria and gives an estimate of the ancient field strength (B$_{anc}$). The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show non-ideal sagging behaviour that fails our acceptance criteria. Specimen BD2-5a is an example where the data appear linear with distinct slopes in the low and high temperature ranges such that it could pass less restrictive selection criteria, particularly if a narrower temperature range was used for the experiment. Data analysis and visualization was conducted using PmagPy \cite{Tauxe2016a}}}
\label{fig:IZZI_examples}
\end{figure*}

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\footnotesize{A) Summary plot of individual specimen absolute paleointensity results (square symbols) and their averages and standard deviations at site level (black bars with grey 1$\sigma$ uncertainty boxes) where each `AX' site is an individual anorthosite xenolith within the Beaver River diabase. All results are corrected for cooling rate with a factor of 0.74. The sites with successful experiments come from 3 regions (Carlton Peak, Silver Bay, Kennedy Creek) which would have cooled at distinct times yielding similar estimates within the each region with differences between regions. B) Regional means calculated from the specimen-level data are compared to the distribution of intensities calculated from a paleomagnetic axial dipole moment model for the past 2 million years (PADM2M; \citealp{Ziegler2011a}) at the latitude corresponding to the paleolatitude of study region (22\textdegree N). C) The mean of the 3 regional means is compared to means calculated from 3 random values drawn from the PADM2M model \cite{Ziegler2011a}. The distribution represents a total of 10,000 iterations of taking 3 random draws and calculating the mean. These comparisons highlight that these paleointensity values are strong in relation to the geomagnetic field over the past 2 million years.}}
\label{fig:PINT_cooling_corrected}
\end{figure*}

@@

\subsection*{Rock magnetism}
\subsubsection*{Coercivity}
Additional rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite contain (titano)magnetite as evidenced through the presence of the Verwey transition (Fig. S2;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S2), indicating that magnetite grains in the specimens either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}. Another difference is that samples which pass paleointensity selection criteria have distinctly higher average median destructive field values than other anorthosite and diabase specimens (Fig. \ref{fig:coercivity}). Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity around 80 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that a population of magnetic grains with more multidomain-like behavior is responsible for the non-ideal paleointensity behaviors during experiments on such specimens \cite{Xu2004a}.

@@

\subsection*{Considering secular variation and cooling rate}
To best characterize the geomagnetic axial dipole field intensity during a certain time period a paleointensity dataset should cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged. Thermal modeling results from ref. \citealp{Zhang2021b} indicate that the Beaver River anorthosite xenoliths were heated to tholeiitic magma temperature ($\sim$1100\textdegree C) and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a thousand years, partially averaging secular variation within single sites. Another consequence of the anorthosite xenoliths having cooled in the interior of thick diabase intrusions is that slow cooling rates can bias paleointensity estimates toward higher values. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref.  \citealp{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully blocked in their characteristic natural remanence magnetization ($\sim$500\textdegree C;  \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Halgedahl1980a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

@@

The other data type that can provide insight into the long-term history of the core's thermal regime and geodynamo is paleomagnetic data---both paleodirectional data that indicate the presence of a geomagnetic field and paleointensity data that constrain the field's strength. Inner core nucleation would have increased the power to the geodynamo which has the potential to manifest as an increase in Earth's surface field \cite{Davies2021a}. An approach combining dynamo simulations and theoretical scaling relationships has predicted that progressive decay of the field's dipole moment would be followed by a rapid increase in geomagnetic field intensity soon after the onset of inner core nucleation such that there is a minimum in dipole moment just before inner core nucleation \cite{Davies2021a}. Other scenarios are possible, however, such as the model-based prediction that while power increases associated with inner core nucleation strengthen Earth's internal magnetic field, that the dynamo becomes more deeply seated in the core diminishes the increase in magnetic field strength at Earth's surface \cite{Aubert2009a, Landeau2017a}. Such a scenario where the dynamo shifts to a greater depth associated with inner core nucleation led \cite{Aubert2009a} to conclude that the increase in power to the dynamo would be difficult to detect with paleointensity data. Ultimately, further observational paleomagnetic records is key as they hold the potential for testing different model predictions and identifying transitions in ancient field strength \cite{Biggin2015a, Bono2019a}.

>>>>

\affil[a]{Department of Earth and Planetary Science, University of California, Berkeley, CA, 94720}
\affil[b]{U.S. Geological Survey, Moffett Field, CA, 94035}
\affil[c]{Department of Earth and Planetary Sciences, Harvard University, Cambridge, MA, 02138}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy due to the exclusion of light elements \cite{Buffett2000a}. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago \cite{Selkin2007a, Biggin2011a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimated thermal conductivity values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, others have implied higher thermal conductivity values and an age for the inner core that is less than 1.5 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b, Frost2022a}, with some suggesting even younger ages ($<$700 Ma;  \citealp{Labrosse2015a, Ohta2016a, Pozzo2022a}). The possibility of late inner core nucleation has motivated proposals of novel power sources to sustain the geomagnetic field through early Earth history including precipitation of light materials such as MgO \cite{Badro2016a, ORourke2016a, ORourke2016b} and SiO$_2$ \cite{Mittal2020a} at the core-mantle boundary. Given that estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of Earth's core and mantle.

@@

Determinations of the absolute value of ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain a record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluids. However, intraplate magmatism associated with large igneous provinces is typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift (Fig. \ref{fig:Geologic_map}) is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma during which there were pulsed intervals of more rapid activity \cite{Swanson-Hysell2021a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior far from the continental margin and subsequent orogenesis. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and paleomagnetic data from rift rocks forms a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it enables an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

@@

In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud magnetic inclusions from alteration. In addition, the weathering of the plagioclase crystals do not resulting in the formation of secondary iron oxides, whereas other minerals such as olivine and pyroxene are more likely to alter and influence the magnetic mineralogy. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They likely cooled from tholeiitic magma temperatures and acquired their magnetizations in conjunction with the host diabase at a paleolatitude of 22 $\pm$ 2\textdegree (calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a}). The details regarding the thermal history of these anorthosite xenoliths are discussed in \cite{Zhang2021b}. Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. The anorthosite xenoliths show low anisotropy of thermal remanent magnetization (TRM) acquisition and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows that the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals without strong preferred orientations. These single-slope, high-quality paleointensity data, together with other high-quality paleointensity records during the protracted Midcontinent Rift magmatic activity, require that there was a strong late Mesoproterozoic geodynamo.

@@

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths to have dominantly single-component magnetization that often unblocks sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with magnetization held by low-titanium titanomagnetite \cite{Zhang2021b}.

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). Plagioclase crystals within coarse-grained intrusions often contain abundant elongate Fe-Ti oxide inclusions that are visible through optical microscopy \cite{Feinberg2006a, Wenk2011a, Ageeva2016a}. The Beaver River anorthosite xenoliths lack such large (10s of $\mu$m) oxide inclusions such that oxides are not visible within the plagioclase crystals through optical microscopy (Fig. \ref{fig:Petro_QDM}C). However, magnetic imaging using a quantum diamond microscope (QDM) shows that there are magnetic remanence carriers within the plagioclase crystals (Fig. \ref{fig:Petro_QDM}E, G). The sources of these magnetic remanence carriers are likely to be Fe-Ti oxides that exsolved within plagioclase crystals above the Curie temperature of magnetite \cite{Zhang2021b, Bian2021a}. To highlight these distinct aspects of the Beaver River anorthosite xenoliths, we present petrographic and magnetic imaging data from the Beaver River anorthosite as well as a sample of an anorthosite from the Duluth Complex Anorthositic Series---an older intrusive complex within the Midcontinent Rift that was not targeted for paleointensity experiments in this study (Fig. \ref{fig:Petro_QDM}). In contrast to the Beaver River anorthosite, the plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display pronounced igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles within the Duluth Complex plagioclase grains that are typically aligned with the [001] axes of the crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these magnetite needles have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}F, H). As a result of this shape anisotropy, magnetization is acquired at angles highly oblique to applied fields (Figure \ref{fig:Petro_QDM}F). In an experiment where a second orthogonal field was applied, the first isothermal remanent magnetizations (IRM) of a set of magnetite needles within a plagioclase grain either fail to rotate or flip by 180\textdegree. When they flip, they remain in a direction that is oblique to the applied field direction dictated by shape anisotropy (Fig. \ref{fig:Petro_QDM}F, H). These experiments enable novel visualization of the grain-scale magnetic anisotropy of elongated exsolved (titano)magnetite inclusions within plagioclase that leads to magnetic anisotropy observed at the bulk sample scale \cite{Selkin2000a, Feinberg2006a}. In contrast, the remanent magnetizations of the ferromagnetic grains imaged in the Beaver River anorthosite xenoliths, targeted for paleointensity in this study, align with the applied field directions indicating minimal remanence anisotropy  (Fig. \ref{fig:Petro_QDM}E, G). The relative lack of petrofabrics and minimal grain-scale magnetic anisotropy make the Beaver River anorthosite xenoliths a particularly compelling target for paleointensity experiments.

@@

Following IZZI-style paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase sites have specimen results that pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1) where each site is an individual anorthosite xenolith. The paleointensity quality index (Q$_{PI}$; \citealp{Biggin2014a}) for the anorthosite xenolith sites are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from the anorthosite xenoliths have a mean of 38.28$\pm$11.92 $\mu$T. The site-mean virtual dipole moment is $\sim$82 ZAm$^2$ ($10^{21} Am^2$) ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthref.org/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments are displayed on Arai plots and zero-field heating results are shown on inset orthogonal plots (Zijderveld plots) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent magnetization (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Dashed red and blue lines show bounding regions calculated for determine the \textit{SCAT} statistic \cite{Shaar2013a}. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent fits for the dominant single-slope component that passes the acceptance criteria and gives an estimate of the ancient field strength (B$_{anc}$). The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show non-ideal sagging behaviour that fails our acceptance criteria. Specimen BD2-5a is an example where the data appear linear with distinct slopes in the low and high temperature ranges such that it could pass less restrictive selection criteria, particularly if a narrower temperature range was used for the experiment. Data analysis and visualization was conducted using PmagPy \cite{Tauxe2016a}}}
\label{fig:IZZI_examples}
\end{figure*}

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\footnotesize{A) Summary plot of individual specimen absolute paleointensity results (square symbols) and their averages and standard deviations at site level (black bars with grey one standard deviation uncertainty boxes) where each `AX' site is an individual anorthosite xenolith within the Beaver River diabase. All results are corrected for cooling rate with a factor of 0.74. The sites with successful experiments come from 3 regions (Carlton Peak, Silver Bay, Kennedy Creek) which would have cooled at distinct times yielding similar estimates within the each region with differences between regions. B) Regional means calculated from the specimen-level data are compared to the distribution of intensities calculated from a paleomagnetic axial dipole moment model for the past 2 million years (PADM2M; \citealp{Ziegler2011a}) at the latitude corresponding to the paleolatitude of study region (22\textdegree N). C) The mean of the 3 regional means is compared to means calculated from 3 random values drawn from the PADM2M model \cite{Ziegler2011a}. The distribution represents a total of 10,000 iterations of taking 3 random draws and calculating the mean. These comparisons highlight that these paleointensity values are strong in relation to the geomagnetic field over the past 2 million years. All shaded regions in B) and C) represent one standard deviation uncertainties.}}
\label{fig:PINT_cooling_corrected}
\end{figure*}

@@

\subsection*{Rock magnetism}
\subsubsection*{Coercivity}
Additional rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite contain (titano)magnetite as evidenced through the presence of the Verwey transition (Fig. S2;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S2), indicating that magnetite grains in the specimens either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}. Another difference is that samples which pass paleointensity selection criteria have distinctly higher average median destructive field values than other anorthosite and diabase specimens (Fig. \ref{fig:coercivity}). Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivities ranging from $\sim$44-144 mT with a median of 53 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that a population of magnetic grains with more multidomain-like behavior is responsible for the non-ideal paleointensity behaviors during experiments on such specimens \cite{Xu2004a}.

@@

\subsection*{Considering secular variation and cooling rate}
To best characterize the geomagnetic axial dipole field intensity during a certain time period a paleointensity dataset should cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged. Thermal modeling results from ref. \citealp{Zhang2021b} indicate that the Beaver River anorthosite xenoliths were heated to tholeiitic magma temperature ($\sim$1100\textdegree C) and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a thousand years, partially averaging secular variation within single sites. Another consequence of the anorthosite xenoliths having cooled in the interior of thick diabase intrusions is that slow cooling rates can bias paleointensity estimates toward higher values. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref.  \citealp{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they blocked in the majority of their characteristic natural remanence magnetization (lab unblocking temperatures $\sim$500\textdegree C;  \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$10 kyr considering the effect of slow cooling on lowering the natural remanence unblocking temperatures \cite{Halgedahl1980a}. This range of cooling time corresponds to a cooling rate of $\sim2.5\times10^{-10}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Halgedahl1980a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. This factor may result in an over-correction, as some multidomain grains that do not tend to severely bias paleointensity estimates toward higher \cite{Biggin2013a}, may exist in the anorthosite xenoliths as is evidenced through the slightly zigzagging Arai plots (Figure \ref{fig:IZZI_examples}). The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

@@

The other data type that can provide insight into the long-term history of the core's thermal regime and geodynamo is paleomagnetic data---both paleodirectional data that indicate the presence of a geomagnetic field and paleointensity data that constrain the field's strength. Inner core nucleation would have increased the power to the geodynamo which has the potential to manifest as an increase in Earth's surface field \cite{Davies2021a}. An approach combining dynamo simulations and theoretical scaling relationships has predicted that progressive decay of the field's dipole moment would be followed by a rapid increase in geomagnetic field intensity soon after the onset of inner core nucleation such that there is a minimum in dipole moment just before inner core nucleation \cite{Davies2021a}. Other scenarios are possible, however, such as the model-based prediction that while power increases associated with inner core nucleation strengthen Earth's internal magnetic field, that the dynamo becomes more deeply seated in the core diminishes the increase in magnetic field strength at Earth's surface \cite{Aubert2009a, Landeau2017a}. Such a scenario where the dynamo shifts to a greater depth associated with inner core nucleation led \cite{Aubert2009a} to conclude that the increase in power to the dynamo would be difficult to detect with paleointensity data. Ultimately, further observational paleomagnetic records are key as they hold the potential for testing different model predictions and identifying transitions in ancient field strength \cite{Biggin2015a, Bono2019a}.
